---
title: "Introduction to Regression - Week 1 Notes"
author: "Trenton Potgieter"
date: "Monday, April 06, 2015"
output: 
    pdf_document:
        toc: true
        latex_engine: xelatex
    fontsize: 11pt
    geometry: margin=1in
---

\newpage

# Francis Galton

Data first used by Francis Galton, who created the terms __Regression__ and __Correlation__ in 1885. And by making use of __Rgression__, we are provided with very interpretable models.

```{r galton, echo = TRUE}
require(UsingR)

#Load the Data
data(galton)

#Plot the Child and parent data
par(mfrow = c(1, 2))
hist(galton$child, col = "blue", breaks = 100) #100 histogram breaks
hist(galton$parent, col = "blue", breaks = 100) #100 histogram breaks

```

The plots above do not descibe the joint relationship. To uderstand the joint relationship we need to first understand summarizing the __marginal__. The marginal is the distribution (on the histograms) of __Children__, disregarding __Parents__ and the distibution of __Parents__, disregarding __Children__, so summarizing the marginal informaion in each Histogram by themselves is a way of describing the "middle" of these datasets.

To do this, let's consider the __children's heights__. So let $Y_i$ be the height of a particular __child__ $i$ for $i=1,\ldots,n$, where $n=928$ (the number of __Children__).

So to define the middle we look for the value of $\mu$ that mimimizes $$\sum_{i=1}^n(Y_i -\mu)^2$$

__The sum of the squared distances between the data and the "middle" value.__

This turns out to be the center of mass of the histogram, the point the mimimizes the average squared distance from all the other points (Least Squares). So in this case the answer is the __sample mean__ ($\mu=\bar{X}$).

## Proof 

To prove this we will use the `manipulate()` function in `R` to se what value of $\mu$ minimizes the sum of squared deviations.

```{r manipulate, echo = TRUE}
require(manipulate)

#Create the Hist() Function
Hist <- function(mu){        
        #Create a histogram of the child data as before
        hist(galton$child, col = "blue", breaks = 100)
        #Draw the line that can be used by manipulate
        lines(c(mu, mu), c(0, 150), col = "red", lwd = 5)
        #Clculate the mean squared error
        mse <- mean((galton$child - mu)^2)
        #Create the labels
        text(63, 150, paste("mu = ", mu))
        text(63, 140, paste("MSE = ", round(mse, 2)))
}

#To call this and use manipulate in R, simply run the following:
#manipulate(Hist(mu), mu = slider(62, 74, step = 0.5))

```


