Using Machine Learning to Predict Weight Lifting Excercises
=========================================================================
## Abstract
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it.

The goal of this report is to predict - by applyting various Machine Learning techniques on data from accelerometers  located on the belt, forearm, arm, and dumbell - the manner in which 6 participants performed the exercises.

__Note:__ More information is available from the website.[^website]

### Data Processing
#### Downloading the data
To start the process, the data[^training] for analysing, testing and building the Machine Learning model, as well as the data[^testing] for testing is downloaded and read into R.

```{r Initialize, echo = TRUE}
## Initialize the environment
if (!file.exists("data")) {
        dir.create("data")
}

## Download the Data
trainURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(trainURL, "./data/train.csv", method = "curl")
download.file(testURL, "./data/test.csv", method = "curl")
downloadDate <- date()

## Read the file into R
trainData <- read.csv("./data/train.csv", header = TRUE)
testData <- read.csv("./data/test.csv", header = TRUE)

```

__Note:__ The data for this report was downloaded on __`r downloadDate`__.

#### Initial Analysis
A first look at the data shows that there are __`r dim(trainData)[[1]]`__ Observastions and __`r dim(trainData)[[2]]`__ variables that can be used as predictors for the Machine Learning algorithm. Since the data set is quite large, we take a random sample of 8 of the features and analyze some of their characteristics.

```{r Sample, echo = TRUE}
# Select 8 random predictors
set.seed(6969)
numPredictors <- ncol(trainData) - 1
s <- sample(1:numPredictors, 8)
Sample <- trainData[, s]
str(Sample)
summary(Sample)

```
As can be seen from the summary of the 8 random samples, there are a lot of missing values as well as some "other" variables. The following plot show the extent of these missing values, in most casses taking entire variables have missing data.

```{r MissingVarPlot, echo = TRUE}
## Plot the spread of missing values in the Sample
missingVar <- sapply(Sample, function(x) sum(is.na(x)))
dfmissing <- data.frame(variable = names(missingVar), missing = missingVar,
                        stringsAsFactors = FALSE)
dfmissing$variable <- factor(dfmissing$variable, levels = dfmissing$variable,
                             ordered =  FALSE)
qplot(x = variable, y = missing, data = dfmissing, geom = "bar", 
      stat = "identity", position = "dodge") +  
        xlab("Preictor Variable") + ylab("No. Missing Values") + 
        theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### Data Transformation and Pre-processing
To clean the data and prepare it for modeling, the first step is to remove the variables that do not pertain to 

#### Cleaning the Data

#### Pre-Processing the Data

### Model Fitting

```{r FitModel, echo = TRUE, cache = TRUE}

```


### Results
#### Pre-process the Testing Data

#### Final Model Test

### Conclusion

### References
Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6.

[^website]: http://groupware.les.inf.puc-rio.br/har
[^training]: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
[^testing]: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv