---
title: "Datacamp data.table Class Exercises"
author: "Trenton Potgieter"
date: "Sunday, November 02, 2014"
output: pdf_document
---

##Chapter 1
###Exercise 1 - Create and Subset a __data.table__
####Instructions

* Create a data.table __my_first_data_table__ with a column x = c("a","b","c","d","e") and a column y = c(1,2,3,4,5). Use data.table().  
* Create a two-column data.table called DT that contains the four integers 1,2,1,2 in the first column a and the letters A,B,C,D in the second column b. Use recycling.  
* Select the third row of DT and just print the result to the console.  
* Select the second and third rows without using any comma at all and print the result to the console.  

```{r exercise_1_1, echo = TRUE}
library("data.table")

# Create your first data.table 
my_first_data_table <- data.table(x = c("a","b","c","d","e"),y = c(1,2,3,4,5))  

# Create a data.table using recycling
DT <- data.table(a = c(1L,2L), ## Note the "L" for integers
                 b = LETTERS[1:4]) ## Note "LETTERs" so no need to do this:
## b = c("A", "B", "C", "D")

# Print the third row to the console
DT[3]

# Print the second and third row to the console, but do not use any comma at all
DT[2:3]

```

###Exercise 2 - Get to know __data.table__
__Remember__ that the special symbol __.N__ contains the number of rows. You can put this symbol inside square brackets

####Instructions
* Select the penultimate row of the table. Make use of .N for this.  
* Return the column names of the data.table.  
* Return the number of rows and number of columns of the data.table.  
* Select row 2 twice and row 3, returning a data.table with three rows where row 2 is a duplicate of row 1.  

```{r exercise_1_2, echo = TRUE}
# Print the penultimate row of DT using `.N`
DT[.N-1]

# Print the column names of DT, and number of rows and number of columns
names(DT) ## or colnames(DT)
dim(DT)

# Select row 2 twice and row 3, returning a data.table with three rows where row 2 is a duplicate of row 1.
DT[c(2,2,3)]

```

###Exercise 3 - Subsetting data tables
__Remember__ the phrase "Take DT, subset rows using __i__, then calculate __j__ grouped by __by__". Also, __j__ can be used to select columns by wrapping the column names in __.()__. In addition to selecting columns, you can also call functions on them as if the columns were variables.

####Instructions
* Create a subset columns B and C for rows 1 and 3, and print the result to the console.
* Assign to ans a data.table that contains two columns: B and val, where val is the product of A and C.
* Assign to answ another data.table answ <- DT[, .(B, val = __BLANK__ )]. Fill in the blanks such that answ equals data.table(B=c("a", "b", "c", "d", "e", "a", "b", "c", "d", "e"), val = as.integer(c(6,7,8,9,10,1,2,3,4,5))).

```{r exercise_1_3, echo = TRUE}
# Create the data,table DT
DT <- data.table(A = 1:5, B = c("a", "b", "c", "d", "e"), C = 6:10)

# Subset rows 1 and 3, and columns B and C
DT[c(1, 3), .(B, C)]

# Assign to ans the correct value
ans <- data.table(B = DT[, B], val = DT[,A * C])
# Note: Official solution is ans <- DT[, .(B, val = A * C)]

# Fill in the blank
answ <- DT[, .(B, val = c(C, A))]

```

###Exercise 4 - The __by__ basics
__Remember__ if you supply a __j__ expression and a by list of expressions, the __j__ expression is repeated for each __by__ group.

####Instructions
* Convert the iris data set to a data.table called DT.
* For each Species, what is the mean Sepal.Length?
* Do exactly the same as in the instruction above, but this time group by the first letter of the Species name instead.

```{r exercise_1_4, echo = TRUE}
# load the iris dataset
data(iris)

# iris as data.table
DT <- as.data.table(iris)

# mean `Sepal.Length`
DT[, mean(Sepal.Length), by = .(Species)]

# Group by the first letter
DT[, mean(Sepal.Length), by = substring(Species, 1, 1)]
# Note: substring(x, start, stop) extracts or replaces the substrings in a
# character vector.

```

###Exercise 5 - Using __.N__ and __by__
__Remember__ that __.N__ can be used in __i__ and that it designates the number of rows in a data.table. There, it is typically used for returning the last row or an offset from it. __.N__ can be used in __j__ too and designates the number of rows in this group. The latter is especially powerful when you can use it in combination with by.

####Instructions
* Group the specimens by Sepal area (to the nearest 10 cm2) and count how many occur in each group.
* Using the answer to the above question, you can now name the group Area and the count Count.

```{r exercise_1_5, echo = TRUE}
# Use the data.table DT for the iris dataset that was used in Exercise 4.

# Group the specimens by Sepal area (to the nearest 10 cm2) and count how many
# occur in each group. For example: DT[,___, by=10*___(___*___ / 10)]
DT[, .N, by = 10 * round(Sepal.Length * Sepal.Width/10)]

# Now name the output columns `Area` and `Count`
DT[, .(Count = .N), by = .(Area = 10 * round(Sepal.Length * Sepal.Width/10))]

```

###Exercise 6 - Return multiple numbers in __j__.
In the previous exercises we returned single numbers in __j__. However, this is not necessary, because you do not have to return only single numbers in __j__.

####Instructions
* Given a new DT, calculate cumulative sum of C in column C while you group by A,B. Store it in a new data.table DT2 with 3 columns A,B and C.
* Select from DT2, the last two values of C in column C while you group by A alone. Use default output names.

```{r exercise_1_6, echo = TRUE}
# Create a new data.table called DT
set.seed(1L)
DT <- data.table(A = rep(letters[2:1], each = 4L), B = rep(1:4, each = 2L),
                 C = sample(8))
DT

# DT2
DT2 <- DT[, .(C = cumsum(C)), by = .(A, B)]
# Note: cumsum(x) returns a vector whose elements are the cumulative sums of
# the elements of the argument

# Now the last two values of C from DT2 while you group by A
DT2[, .(C = tail(C, 2)), by = A] 

```

##Chapter 2
###Exercise 1 - Create and Subset a __data.table__
__Remember__ that chaining allows concatenating multiple operations and follows because the operations are in the same order. Furthermore, it helps to avoid the creation of unnecessary temporary variables (which could quickly clutter one's work space).

####Instructions
* In the Chapter 1, Exercise 6, we calculated DT2 by taking the cumulative sum of C while grouping by A,B. Next, you selected from DT2 the last two values of C while grouping by A alone. Do this again, but this time use the concept of chaining instead.

```{r exercise_2_1, echo = TRUE}
# DT
set.seed(1L)
DT <- data.table(A = rep(letters[2:1], each = 4L), B = rep(1:4, each = 2L),
                 C = sample(8)) 

# Previous exercise:
# DT2 <- DT[, .(C = cumsum(C)), by = .(A, B)]
# DT2[, .(C = tail(C, 2)), by = A]
# Now use chaining
DT[,.(C = cumsum(C)), by = .(A, B)][, .(C = tail(C,2)), by = A]

```

###Exercise 2 - Chaining the __iris__ dataset
Once the __iris__ data set is converted to a data.table called DT. Let us see how you can use chaining to simplify manipulations and calculations with the DT.

####Instructions
* Get the median of all the four columns Sepal.Length, Sepal.Width, Petal.Length and Petal.Width while you group by Species. Give them the same names. Next, order Species in descending order using chaining.

__NOTE:__ This is deliberately difficult and a little bit repetitive. Also errors can easily be made, especially if there are a large amount of columns in the data set. See the next exercise for a simpler solution.

```{r exercise_2_2, echo = TRUE}
# load the iris dataset
data(iris)

# iris as data.table
DT <- as.data.table(iris)

# iris
DT[, .(Sepal.Length = median(Sepal.Length), Sepal.Width = median(Sepal.Width),
       Petal.Length = median(Petal.Length), Petal.Width = median(Petal.Width)),
   by = Species][order(-Species)]
# Note the "-" before Species to denote descending order.

```

###Exercise 3 - Progamming time vs. readability
__Remember__ the data.table package provides a special in-built variable __.SD__. It refers to the subset of data for each group. Additionally, it is a good idea to make use of familiar functions from base R to reduce programming time without losing readability.  

####Instructions
* Get the mean of columns y and z grouped by x by using .SD.
* Get the median of columns y and z grouped by x by using .SD.

```{r exercise_2_3, echo = TRUE}
# Create a sample data.table
DT <- data.table(x = c(2, 1, 2, 1, 2, 2, 1), y = c(1, 3, 5, 7, 9, 11, 13), 
                 z = c(2, 4, 6, 8, 10, 12, 14))
DT

# Mean of columns
DT[, lapply(.SD, mean), by = x]

# Median of columns
DT[, lapply(.SD, median), by = x]

```

###Exercise 4 - Introducing __.SDcols__
__.SDcols__ specifies the columns of a __data.table__ that are included in __.SD__. Using __.SDcols__ comes in handy if you have too many columns and you want to perform a particular operation on a particular subset of the columns (instead from the grouping variable columns). Look at __?data.table__ for more info on __.SDcols__.

####Instructions
* Calculate the sum of the Q columns using .SD and .SDcols. Use 2:4.
* You can of course set .SDcols to be the result of a function call. This time calculate the sum of columns H1 and H2 using paste0().
* Finally, select all but the first row of groups 1 and 2, returning only the grp column and the Q columns. Use -1 in i of .SD and use paste0() again. Type desired_result to better understand what you need to do.

```{r exercise_2_4, echo = TRUE}
# Create a sample data.table
DT <- data.table(grp = c(6, 6, 8, 8, 8), Q1 = c(2, 2, 3, 5, 2),
                 Q2 = c(5, 5, 4, 4, 1), Q3 = c(2, 1, 4, 2, 4),
                 H1 = c(3, 4, 5, 2, 4), H2 = c(5, 2, 4, 1, 2))
DT

# Calculate the sum of the Q columns
DT[, lapply(.SD, sum), .SDcols = 2:4]

# Calculate the sum of columns H1 and H2 
DT[, lapply(.SD, sum), .SDcols = paste0("H", 1:2)]
# Note: paste() converts its arguments to character strings and concatenates
# them (separating them by the string given by sep). paste0() is equivalent to
# paste(..., sep = "", collapse), 

# Select all but the first row of groups 1 and 2, returning only the grp column and the Q columns. 
DT[, .SD[-1], by = grp, .SDcols = paste0("Q", 1:3)]

```

###Exercise 5 - Putting it together: __lapply__, __.SD__, __.SDcols__, __.N__
__Remember__ that __lapply()__ and __.SD__ return a list, and that __.N__ is an inbuilt variable that returns an integer vector of length 1. If __j__ returns a list then a data.table is returned. To combine items together you already know to use c(). Combining a list with a vector makes a new list one longer. When you select __.N__ on its own, it automatically gets named N in the output for convenience when chaining.

####Instructions
* Get the sum of all columns x, y and z and the number of rows in each group while grouping by x. Your answer should be identical to the result stored in desired_result_1.
* Get the cumulative sum of column x and y while grouping by x and z > 8 such that the answer looks like the result that was stored in the variable desired_result_2.
* Use chaining to get the maximum of both x and y by combining the result that was obtained in the previous instruction and group by by1.

```{r exercise_2_5, echo = TRUE}
# Manaually pre-load the sample data
DT <- data.table(x = c(2, 1, 2, 1, 2, 2, 1), y = c(1, 3, 5, 7, 9, 11, 13),
                 z = c(2, 4, 6, 8, 10, 12, 14))
DT

# Sum of all columns and the number of rows
# desired_result_1
#   x x  y  z N
#1: 2 8 26 30 4
#2: 1 3 23 26 3
DT[, c(lapply(.SD, sum), .N), by = x, .SDcols = c("x", "y", "z")]

# Cumulative sum of column `x` and `y` while grouping by `x` and `z > 8`
# desired_result_2
#   by1   by2 x  y
#1:   2 FALSE 2  1
#2:   2 FALSE 4  6
#3:   1 FALSE 1  3
#4:   1 FALSE 2 10
#5:   2  TRUE 2  9
#6:   2  TRUE 4 20
#7:   1  TRUE 1 13
DT[, lapply(.SD, cumsum), by = .(by1 = x, by2 = z > 8), .SDcols = c("x", "y")]

# Chaining
DT[, lapply(.SD, cumsum), by = .(by1 = x, by2 = z > 8), .SDcols = 1:2][, lapply(.SD, max), by = by1, .SDcols = c("x", "y")]

```

###Exercise 6 - Adding, Updating and removing columns (LHS := RHS)
__Remember__ that __:=__ is defined for use in __j__ only, and there are two ways of using it.  
The first is the __LHS := RHS__ form, where __LHS__ is a character vector of column names and __RHS__ is the corresponding value.

####Instructions
* Add a column Total by reference containing sum(B) for each group in column A.
* Add 1 to column B just in rows 2 and 4.
* Add a new column Total2 that contains sum(B) grouped by A but just over rows 2,3 and 4.
* Remove the Total column. The R function [[ is useful to select a single column by name or number with the result returned as a vector. Use [[ to select the third column.

__NOTE:__ that for the second instruction in j the performance goes up if you coerce RHS to integer yourself via 1L or via as.integer(). 

```{r exercise_2_6, echo = TRUE}
# Manaually pre-load the sample data
DT <- data.table(A = c("a", "a", "a", "b", "b"), B = 1:5)
DT

# Column `Total` by reference
DT[, Total := sum(B), by = A]
DT

# Add 1 to column `B`
DT[c(2, 4), B := B + 1L]
DT

# Add a new column `Total2`
DT[2:4, Total2 := sum(B), by = A]
DT

# Remove the `Total` column
DT[, Total := NULL]
DT

# Select the third column using `[[`
DT[[3]]

```

###Exercise 7 - The Function form of :=
The second way to use __:=__ is a function form. The nice thing about a function form is that you can get each right hand side (RHS) alongside the left hand side (LHS) so it is easier to read.

####Instructions
* Update B with B+1 and add a new column C with A+B, and D with constant 2.
* Given myCols = c("B","C"), delete those columns.
* Delete column D by its number (2) rather than its name (D).

```{r exercise_2_7, echo = TRUE}
# Manaually pre-load the sample data
DT <- data.table(A=c(1,1,1,2,2), B=1:5)
DT

# Update `B`, add `C` and `D`
DT[, `:=` (B = B + 1, C = A + B, D = 2)]
DT

# Delete `myCols`
myCols = c("B","C")
DT[, (myCols) := NULL]
DT

# Delete column 2 by number
DT[, 2 := NULL]
DT

```

###Exercise 8 - Using set().
__Remember__ that __set()__ can not do grouped operations.

####Instructions
* Loop through columns 2,3 and 4, and for each one select 3 rows at random and set the value of that column to NA.
* Change the column names to lower case. When setnames() is passed just one input vector, that needs to be all the new names.

```{r exercise_2_8, echo = TRUE}
set.seed(1)

# Manually create DT
DT <- data.table(A = c(2, 2, 3, 5, 2, 5, 5, 4, 4, 1),
                 B = c(2, 1, 4, 2, 4, 3, 4, 5, 2, 4),
                 C = c(5, 2, 4, 1, 2, 2, 1, 2, 5, 2),
                 D = c(3, 3, 3, 1, 5, 4, 4, 1, 4, 3))
DT

# For loop with set
for(i in 2:4)
    set(DT, sample(10, 3), i, NA)

# Change the column names to lowercase
setnames(DT, tolower(names(DT)))

# Print the new DT to the console to confirm changes
DT

```

###Exercise 9 - The set() family
__Remember__ the following:  
1. set() is a loopable low overhead version of ":="  
2. You can use setnames() to set or change column names  
3. setcolorder() lets you reorder the columns of a data.table

####Instructions
* Add a postfix "_2" to all column names.
* Change "a_2" to "A2" by names not position.
* To end, reverse the order of the columns.

```{r exercise_2_9, echo = TRUE}
# Create the sample dataset
DT <- data.table(a=letters[c(1,1,1,2,2)], b=1)
DT

# Postfix "_2"
setnames(DT, paste0(names(DT), "_2"))
DT

# "a_2" to "A2"
setnames(DT, "a_2", "A2")
DT

# Reversing order
# NOTE: the usage of the 2:1
setcolorder(DT, 2:1)
DT

```

##Chapter 3
###Exercise 1 - Selecting rows the data.table way
__Remember__ that you can use column names in __i__ to select certain rows. Since selecting rows over and over again is common place with __data.table__, it is best to do a small exercise on this with the help of the familiar iris data set. 

__NOTE:__ This is different from selecting columns and running functions on columns using __j__.

####Instructions
* Convert the iris data set to a data.table with the same name iris
* Select all the rows where Species is "virginica"
* Select all the rows where Species is either "virginica" or "versicolor"

```{r exercise_3_1, echo = TRUE}
# iris as data.table
iris <- as.data.table(iris) 

# Species is "virginica"
iris[Species == "virginica"]

# Species is either "virginica" or "versicolor"
iris[Species %in% c("virginica", "versicolor")]

```

###Exercise 2 - Removing columns and adapting column names
In the previous exercise you needed to select certain rows out of the iris data.table based on the column names. Now you have to take your understanding of the data.table package to the next level: use standard R functions and regular expressions to manipulate your data.table by removing columns and adapting column names. To practice this, we'll do a little manipulation to prepare easier column names for the next exercise. 

####Instructions  
* Remove the "Sepal." prefix and remove the two Petal columns from the iris data.table.

```{r exercise_3_2, echo = TRUE}
# Convert iris to a data.table
iris <- as.data.table(iris)

# Remove the "Sepal." prefix... 
# Note that gsub() performs replacement of all matches
# of the regular expression
setnames(iris, gsub("^Sepal[.]", "", names(iris)))
iris

# ...and remove the two `Petal` columns
iris[, grep("^Petal", names(iris)) := NULL]
iris

```

###Exercise 3 - Understanding automatic indexing
__Remeber__ the rule that __"if i is a single variable name, it is evaluated in calling scope, otherwise inside DT's scope"__. This is a very important rule if you want to conceptually understand what is going on when using column names in __i__. Only single columns on the left of operators benefit from automatic indexing. 

####Instructions
* Select the rows where the area is greater than 20 square centimeters.
* Add a new boolean column containing Width x Length > 20 and call it IsLarge.
* Select the rows where IsLarge is TRUE.

```{r exercise_3_3, echo = TRUE}
# Use the data.table iris that is already loaded
iris

# Area is greater than 20 square centimeters
iris[Width * Length > 20]

# Now demonstrate the above conceptually
# Add new boolean column
iris[, IsLarge := Width * Length > 20]
iris

# Now select rows again where the area is greater than 20 square centimeters
iris[IsLarge == TRUE]
# Note that iris[(IsLarge)] can also be used

```

###Exercise 4 - Testing the understanding of KEYs
####Instructions
Load the sample data.table (DT) and execute the following:

```{r exercise_3_4, echo = TRUE}
# Load the sample data.table
DT <- data.table(A = c("b", "a", "b", "c", "a", "b", "c"),
                 B = c(2, 4, 1, 7, 5, 3, 6),
                 C = 6:12)

```

1. Select the b group using ==.
```{r exercise_3_4_1, echo = TRUE}
DT[A == "b"]

```
2. Set a 2 column key on A and B.
```{r exercise_3_4_2, echo = TRUE}
setkey(DT, A, B)
```
3. Select the b group again: again use ==.
```{r exercise_3_4_3, echo = TRUE}
DT[A == "b"]

```

__Note__ that the order of the rows within the b group changed, because __B__ is included in the key.  

###Exercise 5 - Selecting groups or parts of groups
The previous exercise illustrated how you can manually set a key via __setkey(DT, A, B)__. __setkey()__ will then sort the data by the columns that you specify, and change the table by reference. Having set a key will allow you to use it as a __super-charged__ rowname when doing selections for example. Arguments like __mult__ and __nomatch__ then help you to only select particular parts of groups.

Furthermore, two of the instructions will require you to make use of __by=.EACHI__. This will allow you to run __j__ for each group in which each item in __i__ joins too.

####Instructions
* Select the b group without using ==.
* Select the b and c groups.
* Select the first row of the b and c groups using mult.
* Use by=.EACHI and .SD to select the first and last row of the b and c groups.
* Extend the previous command to print out the group before returning the first and last row from it.

```{r exercise_3_5, echo = TRUE}
# Create the sample data.table (DT).
# Set the keys to `A` and `B`
DT <- data.table(A = letters[c(2,1,2,3,1,2,3)],
                 B = c(5,4,1,9,8,8,6), C=6:12)
setkey(DT,A,B)

# Select the `b` group
DT["b"]

# `b` and `c` groups
DT[c("b", "c")]

# The first row of the `b` and `c` group
DT[c("b", "c"), mult = "first"]

# `by=.EACHI` and `.SD` 
DT[c("b", "c"), .SD[c(1, .N)], by = .EACHI]

# Print out all the data in the two groups before you return the first and last row of each group again.  Use {} and .N 
DT[c("b", "c"), {print(.SD); .SD[c(1, .N)]}, by = .EACHI]

```

###Exercise 6 - Rolling joinbs (Part one)
__Remember__ that the roll applies to the last join column.

####Instructions
* Get the key of DT via the key() function.
* Use the super charged row names to lookup the row where A=="b" & B==6, without using ==. Verify here that C is NA.
* Base yourself on the query that was written in the previous instruction to return the prevailing row before this gap.
* Similar, but this time find the nearest one.

```{r exercise_3_6, echo = TRUE}
# Load the samplke the data.table `DT`
DT <- data.table(A = letters[c(1, 1, 2, 2, 2, 3, 3)],
                 B = c(4, 8, 1, 5, 8, 6, 9),
                 C = c(7, 10, 8, 6, 11, 12, 9))
setkey(DT, A, B)

# Key of `DT`
key(DT)
# Row where  `A=="b"` & `B==6`
DT[.("b", 6)]

# Return the prevailing row
DT[.("b", 6), roll = TRUE]

# Nearest one
DT[.("b", 6), roll = "nearest"]

```

###Exercise 7 - Rolling joins (Part two)
__Remember__ that the __rollends__ argument is actually a vector of two logical values. If you want to roll for a certain distance, you remain with the roll argument. 

####Instructions
* Look at the sequence (-2):10 for the b group.
* Now carry the prevailing values forwards through the NA.
* Next, carry the first observation backwards as well.
* Finally, only roll for a distance of 2, still with rollends=TRUE. Check that only 4 and -2 now have NA. This is because they are more than 2 from the prevailing observation.

__NOTE:__ The desired result should look as follows:

```{r, echo = TRUE}
#    A  B  C
# 1: b -2 NA
# 2: b -1 NA
# 3: b  0 NA
# 4: b  1  8
# 5: b  2  8
# 6: b  3  8
# 7: b  4 NA
# 8: b  5  6
# 9: b  6  6
#10: b  7  6
#11: b  8 11
#12: b  9 11
#13: b 10 11

```

```{r exercise_3_7, echo = TRUE}
# Load the samplke the data.table `DT`
DT <- data.table(A = letters[c(1, 1, 2, 2, 2, 3, 3)],
                 B = c(4, 8, 1, 5, 8, 6, 9),
                 C = c(7, 10, 8, 6, 11, 12, 9))
setkey(DT, A, B)

# Look at the sequence (-2):10 for the `b` group
DT[.("b", (-2):10)]

# Carry the prevailing values forwards
DT[.("b", (-2):10), roll = TRUE]

# Carry the first observation backwards
DT[.("b", (-2):10), roll = TRUE, rollends = TRUE]

# Note that the following can also be used
# DT[.("b", (-2):10), roll = TRUE, rollends = c(TRUE, TRUE)]

# Roll for a distance of 2
DT[.("b", (-2):10, roll = 2)]

```

###Exercise 8 - Rolling joins (Part three)
If you look up the value B==20 in group A=="b" without limiting the roll, the value of column C is now __11__.

```{r exercise_3_8, echo = TRUE}
DT[.("b", 20), roll = TRUE]

```