---
title: "Datacamp data.table Class Exercises"
author: "Trenton Potgieter"
date: "Sunday, November 02, 2014"
output: pdf_document
---

##Chapter 1
###Section 1
####Exercise 1 - Instructions

* Create a data.table __my_first_data_table__ with a column x = c("a","b","c","d","e") and a column y = c(1,2,3,4,5). Use data.table().  
* Create a two-column data.table DT that contains the four integers 1,2,1,2 in the first column a and the letters A,B,C,D in the second column b. Use recycling.  
* Select the third row of DT and just print the result to the console.  
* Select the second and third rows without using any comma at all and print the result to the console.  

```{r exersize_1, echo = TRUE}
library("data.table")

# Create your first data.table 
my_first_data_table <- data.table(x = c("a","b","c","d","e"),y = c(1,2,3,4,5))  

# Create a data.table using recycling
DT <- data.table(a = c(1L,2L), ## Note the "L" for integers
                 b = LETTERS[1:4]) ## Note "LETTERs" so no need to do this:
## b = c("A", "B", "C", "D")

# Print the third row to the console
DT[3]

# Print the second and third row to the console, but do not use any comma at all
DT[2:3]

```

####Exercise 2 - Instructions
* Select the penultimate row of the table. Make use of .N for this.  
* Return the column names of the data.table.  
* Return the number of rows and number of columns of the data.table.  
* Select row 2 twice and row 3, returning a data.table with three rows where row 2 is a duplicate of row 1.  

```{r exercise_2, echp = TRUE}
# The data.table DT is still loaded in your workspace
# Print the penultimate row of DT using `.N`
DT[.N-1]

# Print the column names of DT, and number of rows and number of columns
names(DT) ## or colnames(DT)
dim(DT)

# Select row 2 twice and row 3, returning a data.table with three rows where row 2 is a duplicate of row 1.
DT[c(2,2,3)]

```
